---
title: "PHW251 Data Project Milestone 4"
author: "Nakisa Golabi, Joanna Liang, Desmond Gill"
date: "11/16/2022"
output: pdf_document
---


\newpage
```{r import,include=F}
library(readr)
library(readxl)
library(tidyverse)
library(tibble)
library(dplyr)

url_file = "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_csc_smoker_data.csv"

smoker_dataset <- read_csv(url_file)

url_file_2 = "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_csc_outcome_race_data.csv"

smoker_dataset_race <- read_csv(url_file_2)

dataset_rename <- rename(smoker_dataset, ID = "psraid")

dataset_rename 
head(dataset_rename)
```

\newpage
```{r, include=F}
#Subset rows or columns

smoker_dataset_clean <- rename_with(smoker_dataset, ~ tolower(gsub(" ","_", .x, 
                                                                 fixed=TRUE)))

smoker_dataset_race_clean <- rename_with(smoker_dataset_race, ~ tolower(gsub(" ","_", .x, 
                                                                 fixed=TRUE)))

smoker_dataset_race_clean$id<-gsub("DIS","",as.character(smoker_dataset_race_clean$id))
smoker_dataset_race_clean$id<-gsub("STAT","",as.character(smoker_dataset_race_clean$id))

dataset1 <- rename(smoker_dataset_clean, id=psraid)

jointdataset <- merge(dataset1, smoker_dataset_race_clean, by = c('id'))

```

\newpage
```{r, include=F}
#Create new variables needed for analysis (minimum 2)

jointdataset_2 <- na_if(jointdataset, "(DO NOT READ) Don't know")
jointdataset_3 <- na_if(jointdataset_2, "(DO NOT READ) Refused")

jointdataset_3$smkage[jointdataset_3$smkage=="5 years old or less"]<-"5"
jointdataset_3$smkage[jointdataset_3$smkage=="Never smoked regularly"]<-"0"
jointdataset_3$smk1age[jointdataset_3$smk1age=="5 years old or less"]<-"5"

jointdataset_4 <- drop_na(jointdataset_3, smk1age, smkage)

jointdataset_4$smkage <- as.numeric(as.character(jointdataset_4$smkage))
jointdataset_4$smk1age <- as.numeric(as.character(jointdataset_4$smk1age))

#Variable 1
jointdataset_5 <- jointdataset_4 %>% 
   mutate(smoker_onset = (smkage - smk1age))

jointdataset_5$howmany[jointdataset_5$howmany=="100 or more cigarettes"]<-"100"
jointdataset_6 <- drop_na(jointdataset_5, howmany)

#Variable 2
clean_data <- jointdataset_6 %>% group_by(smokstat, howmany) %>%
  mutate(smoke_level= case_when(
    smokstat=="Current nondaily smoker" & howmany <= 10 ~ "nondaily low level", 
    smokstat=="Current nondaily smoker" & howmany <= 30  ~ "nondaily medium level", 
    smokstat=="Current nondaily smoker" ~ "nondaily high level", 
    smokstat=="Current daily smoker" & howmany <= 10 ~ "daily low level", 
    smokstat=="Current daily smoker" & howmany <= 30  ~ "daily medium level", 
    smokstat=="Current daily smoker" ~ "daily high level"
  ))

#Variable 3
clean_data2 <- drop_na(clean_data, smok6num)
clean_data3 <- drop_na(clean_data2, smok6uni)

clean_data3$howmany <- as.numeric(as.character(clean_data3$howmany))
clean_data4 <- mutate(clean_data3, packs = howmany/20)

clean_data4$smok6num <- as.numeric(as.character(clean_data4$smok6num))

clean_data5 <- clean_data4 %>% mutate(smoking_years = 
                               case_when(
               smok6uni == "Years" ~ smok6num, 
               smok6uni == "Months" ~ smok6num/12,
               smok6uni == "Days" ~ smok6num/365,
               ))

clean_data6 <- clean_data5 %>% mutate(pack_years= round(packs*smoking_years, 2))
```

\newpage
```{r, include=F}
#Clean variables needed for analysis (minimum 2)

#smkbrand
clean_data6$smkbrand[clean_data6$smkbrand=="Other (SPECIFY)"]<-"Other"
clean_data7 <- drop_na(clean_data6, smkbrand)

#wherebuy
clean_data7$wherebuy[clean_data7$wherebuy=="Somewhere else (SPECIFY)?"]<-"Somewhere else"
clean_data8 <- drop_na(clean_data7, wherebuy)

#not including consider in analysis

#smokmore
clean_data9 <- drop_na(clean_data8, smokmore)

#heartdis
clean_data10 <- drop_na(clean_data9, heartdis)

#othmenill
clean_data11 <- drop_na(clean_data10, othmenill)

#diabetes
clean_data12 <- drop_na(clean_data11, diabetes)

#asthma
clean_data13 <- drop_na(clean_data12, asthma)

#race
clean_data14 <- clean_data13 
clean_data14$race01 <- if_else(clean_data13$race01 =="Yes",1,0)
clean_data14$race02 <- if_else(clean_data13$race02 =="Yes",1,0)
clean_data14$race03 <- if_else(clean_data13$race03 =="Yes",1,0)
clean_data14$race04 <- if_else(clean_data13$race04 =="Yes",1,0)
clean_data14$race05 <- if_else(clean_data13$race05 =="Yes",1,0)
clean_data14$race06 <- if_else(clean_data13$race06 =="Yes",1,0)
clean_data14$race12 <- if_else(clean_data13$race12 =="Yes",1,0)
clean_data14$race07 <- if_else(clean_data13$race07 =="Yes",1,0)
clean_data14$race08 <- if_else(clean_data13$race08 =="Yes",1,0)
clean_data14$race09 <- if_else(clean_data13$race09 =="Yes",1,0)
clean_data14$race10 <- if_else(clean_data13$race10 =="Yes",1,0)
clean_data14$race13 <- if_else(clean_data13$race13 =="Yes",1,0)
clean_data14$race11 <- if_else(clean_data13$race11 =="Yes",1,0)
clean_data14$race14 <- if_else(clean_data13$race14 =="Yes",1,0)
clean_data14$race15 <- if_else(clean_data13$race15 =="Yes",1,0)

clean_data14$race01[is.na(clean_data14$race01)] = 0
clean_data14$race02[is.na(clean_data14$race02)] = 0
clean_data14$race03[is.na(clean_data14$race03)] = 0
clean_data14$race04[is.na(clean_data14$race04)] = 0
clean_data14$race05[is.na(clean_data14$race05)] = 0
clean_data14$race06[is.na(clean_data14$race06)] = 0
clean_data14$race12[is.na(clean_data14$race12)] = 0
clean_data14$race07[is.na(clean_data14$race07)] = 0
clean_data14$race08[is.na(clean_data14$race08)] = 0
clean_data14$race09[is.na(clean_data14$race09)] = 0
clean_data14$race10[is.na(clean_data14$race10)] = 0
clean_data14$race13[is.na(clean_data14$race13)] = 0
clean_data14$race11[is.na(clean_data14$race11)] = 0
clean_data14$race14[is.na(clean_data14$race14)] = 0
clean_data14$race15[is.na(clean_data14$race15)] = 0

sum(clean_data14$race11)
sum(clean_data14$race14)
sum(clean_data14$race15)

clean_data15 <- clean_data14 %>% 
  mutate(race_num = race01+ race02+ race03+ race04+ race05 +race06+
           race12 + race07+ race08 + race09+ race10+ race13)

clean_data16 <- filter(clean_data15, id != 100099 & id != 100109 & id != 100191 
                       & id != 100206 & id != 100232)
      

```


\newpage
PRINT-QUALITY TABLE THAT SHOWS THE AVERAGE NUMBER OF PACK-YEARS AND THE 
DISEASE OUTCOMES 


```{r, include=F}

asthma_data <- filter(clean_data16, asthma=="Yes")
mean(asthma_data$pack_years)
#mean asthma pack_years 23.9261

heartdis_data <- filter(clean_data16, heartdis=="Yes")
mean(heartdis_data$pack_years)
#mean heartdis pack_years 28.82653

othmenill_data <- filter(clean_data16, othmenill=="Yes")
mean(othmenill_data$pack_years)
#mean othmenill pack_years 18.94233

diabetes_data <- filter(clean_data16, diabetes=="Yes")
mean(diabetes_data$pack_years)
#mean diabetes pack_years 30.10132

tibble_py_disease <- tibble(
  disease_outcomes = c("Asthma","Heart Disease","Mental Illness","Diabetes"),
  avg_pack_years = c(23.9261, 28.82653, 18.94233, 30.10132), 
)

library(kableExtra)
library(tidyverse)
library(ggthemes)

```



```{r print quality table for pack-years and disease outcomes}

kable(tibble_py_disease, booktabs = T, digits = 2, escape=F,
                   col.names=c("Outcomes", "Pack-Years"), 
                   align="cc",
                   caption = "Average Pack-Years Per Disease Outcome")


```

The table compares the average number of pack-years (the product of the number 
of packs of cigarettes smoked per day and the years a person has smoked) by 
the disease outcomes of asthma, heart disease, diabetes, and mental illness. 
Those with diabetes had the highest mean and those with mental illness 
had the lowest mean. 


\newpage
PLOT COMPARING RACE AND BEHAVIORAL FACTOR OF CIGARETTE BRANDS PURCHASED
```{r, include=F}

clean_data17 <- clean_data16 %>% mutate_at(c('race_num'), ~na_if(., 0))
clean_data17 <- drop_na(clean_data17, race_num)

df1 <- clean_data17 %>% select(smkbrand, race_num)

df2 <- df1 %>%
  group_by(smkbrand, race_num) %>%
  summarize(counts = n()) %>%
  mutate(race_char=as.character(race_num))


```



```{r plot comparing race and behavioral factor of cigarette brand}

ggplot(data=df2, aes(x = counts, y=factor(smkbrand), fill=race_char)) + 
  geom_bar(aes(fill=race_char), stat="identity", position=position_dodge2()) +
  labs(x="Counts",y="Cigarette Brands Purchased",
       title="Cigarette Brands Purchased vs. Racial Identification",
    caption="Data Source: California Department of Public Health 2011 California Smokers Cohort.")+
  scale_fill_manual(name="Self-Reported Races by Each Smoker",
                    values=c("cyan3","darkorange", "darkgreen"))+
  theme(legend.position="bottom")
 


```

This visualization compares the cigarette brands purchased by each smoker with 
their racial identification (number of races self-reported), in the CDPH 2011 
California Smokers Cohort. Out of 712 total smokers, Marlboro is the highest 
purchased cigarette brand by 290 smokers with self-reported one race and by 14 
smokers with self-reported two races; Camel is the highest purchased cigarette 
brand by 2 smokers with self-reported three races. 

\newpage
PLOT COMPARING RACE AND SMOKER ONSET
```{r, include=F}
df3 <- clean_data17 %>% select(smoker_onset, race_num)
df3 <- df3 %>% mutate(race_char=as.character(race_num))

library(dplyr)
df4 <- df3 %>% mutate(smoker_onset=replace(smoker_onset, which(smoker_onset<0), NA))
df4 <- drop_na(df4, smoker_onset)

df_race1 <- filter(df4, race_char=="1")
df_race2 <- filter(df4, race_char=="2")
df_race3 <- filter(df4, race_char=="3")

median(df_race1$smoker_onset)
median(df_race2$smoker_onset)
median(df_race3$smoker_onset)

```

```{r plot comparing race and smoker onset}
  
ggplot(data=df4, aes(x=race_char,y=smoker_onset)) +
  geom_boxplot(aes(colour=race_char)) +
  theme(legend.position="none") +
  labs(x="Self-Reported Races by Each Smoker",y="Smoker Onset", 
       title="Comparing Smoker Onset with Racial Identification", caption=
  "Data Source: California Department of Public Health 2011 California Smokers Cohort.")


```

This box plot compares smoker onset (the difference between the age when they 
started smoking regularly and the age of the participant's first cigarette) 
with their racial identification (number of races self-reported), in the 
CDPH 2011 California Smokers Cohort. The medians for smoker onset are
2.00, 2.00 and 4.00 years for smokers with self-reported one race, two races, 
and three races, respectively. 






